\documentclass[12pt]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{crimson}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{amsmath}
\usepackage{parskip}

% Page setup
\geometry{margin=1in}
\doublespacing
\setlength{\parskip}{1.5em}
\hyphenpenalty=10000
\exhyphenpenalty=10000
\emergencystretch=3em

% Title information
% \title{Ambient Co-Presence and the Social Imagination of Small Urban Spaces}
\title{Ambient Co-Presence and Social Imagination in Urban Spaces}
\title{\huge Ambient Co-Presence \\ \large and\\ \vspace{-1em} \huge Social Imagination in Small Urban Spaces}
\author{Deniz Aydemir}
\date{\today}

\begin{document}

\maketitle

\subsection{Towards a generalization of people watching}
Most of our social experiences are not social interactions. We frequently see, walk by, sit near, and notice others. But we only interact with a small subset of those we are near. We might call this form of non-interactive proximity \textit{ambient co-presence}.\footnote{The term ``ambient co-presence'' is adapted from \citet{madianou2016} where it is used to describe peripheral awareness of distant others through digital media. We repurpose it here for physical urban contexts.}

In cities, the divergence between those with whom we share ambient co-presence and those with whom we actually interact is especially pronounced. We might rub shoulders with hundreds of people for every one interaction. On a walk down a busy street in New York City, that could become thousands.

Much of existing social theory attempts to describe how our interactions and relationships shape the way we think about community, norms, and belonging. In recent years, there have been more explorations of how ambient or fleeting modes of social experience develop our social conceptions \citep{zahnow2025comfort, blokland2014}. But these explorations underspecify the diversity of ambient social experiences, and don't explain the mechanism for their effects. 

Our goals here are to (1) provide a typology of \textit{ambient co-presence} that can be used to add greater precision to discussions of non-interactive social experiences, and (2) show examples of how our imagination uses ambient co-presence in shared spaces to develop our sense of community and belonging. 

These non-interactive, fleeting social experiences are foundational in city life, and understanding them better may help us create healthier communities in our cities.

\section{A typology of co-presence}

We can describe a moment where co-presence occurs between two individuals as an \textit{event} where the two individuals are in perceptible distance to each other. If either of the two individuals can see, hear, smell, or touch the other we will consider them to be in co-presence. 

We can then divide these events into two types we've already described: co-presence that involves an interaction (maybe we can call this \textit{engaged} co-presence) and co-presence that does not, which we call \textit{ambient}. This division is not always clear. How many words shared constitutes an interaction? Is eye contact an interaction? Is wordlessly holding a door for someone else an interaction?

For our purposes, we should focus only on co-presence that can be uncontroversially defined as non-interactive. This means that the strongest co-presence we will consider ambient is an event where person A consciously observes person B, and person A has no knowledge of whether B reciprocated in any conscious noticing of A. Considering ambient co-presence this way allows us to guarantee that any impact on A from the event is irrespective of B's actions. If A consciously sees B acknowledging or responding to A in any way, then we will not consider that event to be one of ambient co-presence. For example, if A and B make eye contact their interaction is no longer only ambient.

This is likely too strong. When you make room for someone to sit next to you on a bus, you may interact with body language or eye contact. But that still seems to qualify as an ambient event, especially if the majority of your time in proximity is spent sitting next to each other without acknowledgment or any further communication. But we won't deal with such edge cases today. 

The operative concept we will use to qualify ambient co-presence is that the whole event is experienced inside A's own mind, and any impacts on A occur solely in A's imagination. In a purely ambient co-presence event, nothing is enacted or tested outside the mind. B does not offer anything other than their ambient effects on A's senses, no non-ambient form of meaningful communication. Despite this, these ambient encounters do shape the way we think. This power of ambient co-presence is what we are trying to tease out.

Because this ambient co-presence is a subjective experience, we can also differentiate between the observer individual (A) and the object individual (B) of the event. When we speak of the impacts of the experience, we will assume those impacts are only relevant to the observer. Naturally, B can be having a simultaneous ambient co-presence event where B is the observer and A is the object, but because it is not a tangible interaction the two simultaneous events may have no relevance to each other.

\subsection{Conscious and Unconscious}

We have described a one-sided, conscious observation as an example of an ambient co-presence event. But we can also include the unconscious or subconscious noticing of others as well. Ambient co-presence could also be described by B sitting in A's peripheral vision, with no need for A to consciously observe or even acknowledge to themselves the presence of B. This allows us to differentiate two types of ambient co-presence: \textit{conscious} and \textit{unconscious}.

\subsection{Familiar and Unfamiliar}

But this is not the only way we can differentiate types of ambient co-presence. In an ambient co-presence event, the observer may recognize the subject, perhaps on a commuter train where riders might often ride with familiar faces, or at a bar where others also frequent. So we can differentiate between \textit{familiar} and \textit{unfamiliar} ambient co-presence.

Familiarity here needs some explanation. Are repeated, conscious, ambient co-presence events between two individuals still ambient if they recognize each other?

As long as the two participants have not interacted before, we will consider repeated ambient co-presence events to be very much possible. It is natural that familiar strangers would be more likely to interact than unfamiliar strangers, and there are certainly interesting things to be said about the kinds of interactions that familiar strangers might have and the implications of those social interactions. For now, we will only be concerned with the impacts of familiar strangers who do not cross the threshold to engaging each other.

\subsection{Types of ambient co-presence}

% TODO: Create table of types of ambient co-presence
% Table should cross conscious/unconscious with familiar/unfamiliar

\vspace{1em}
\begin{table}[H]
\centering
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{lcc}
\toprule
 & \textbf{Familiar} & \textbf{Unfamiliar} \\
\midrule
\textbf{Conscious} & Recognizing a regular on your commute & Noticing a stranger's outfit \\
\textbf{Unconscious} & Assuming familiar others are near & Peripheral awareness of another's presence \\
\bottomrule
\end{tabular}
\renewcommand{\arraystretch}{1.5}
% \caption{Types of Ambient Co-Presence}
\label{tab:types}
\end{table}


There is no doubt that ambient co-presence is a pervasive part of the human experience. 

Sure, we can imagine small towns with tight-knit communities where almost no co-presence goes unengaged -- everyone acknowledges one another at every opportunity. Perhaps this is the kind of community humans evolved to live in. But this is not what social life looks like in most places today, and certainly not in cities. 

\section{A small world of urban spaces}

In cities, we share many spaces with others we never interact with. Let's describe three examples that we can use to explore the different types of ambient co-presence and their implications.

\subsection{Bea sits in a bar}
Bea decides to grab a drink at a local bar she's been meaning to visit. She sits at the bar, interacts with the bartender, but otherwise keeps to herself. There are some individuals, some couples, and some groups of three or four scattered around the bar. In addition to the bartender, there are two servers who are attending those sitting at tables. 

\subsection{Sue walks down a street}
Sue is walking down a crowded street at the end of the workday. She sees many people walking at various degrees of urgency, many people standing, some shopkeepers, some loitering. She sees people turning onto smaller streets, she sees people who look lost, and she sees people going in and out of stores.

\subsection{Trina commutes on a train}
Trina always takes the commuter rail to work. This morning is no different. She sees the usual crowd, along with some she doesn't recall seeing before. She sits at an empty pair of seats, but a couple of stops later an unfamiliar stranger sits next to her.

\section{Ambient co-presence activates our imagination of urban community}

Ambient co-presence activates our imagination. We don't need to interact with those around us to feel a sense of community and belonging.

We hypothesize here that the mechanism that creates a sense of comfort, belonging, and social cohesion is the imagined connection and camaraderie we create through ambient co-presence events. Familiar ambient co-presence especially supports this hypothesis. We know that recognizing strangers, even those with whom we have no interaction, leads to feelings of security, safety, and comfort \citep{zahnow2024security}. 

This makes sense. Familiarity breeds comfort. But what is the mechanism that converts ambient familiarity into comfort? And is ambient familiarity really sufficient or necessary to foster belonging?

\subsection{With familiar ambient co-presence}

Zahnow and Corcoran see feelings of belonging and safety grow both when observers experience a reciprocation of familiarity (a nod from a stranger) and when they don't. Zahnow and Corcoran believe that recognition creates a reciprocal signal of acceptance in the in-group, but they explain this only by saying that the observer experiences a ``sense of shared, symbolic identification'' \citep{zahnow2024security}. 

So Zahnow and Corcoran posit that the observer's imagination is doing the heavy-lifting of creating a "symbolic identification", but they underemphasize that this feeling of acceptance does not require any acknowledgment from the familiar object individual. Ambient familiarity alone triggers our imagined sense of community.

We can think here of Trina. She sees familiar faces on her train, and she might feel a sense of camaraderie for their shared experience. She may notice when someone who's usually on the commute isn't one day, even if they've never interacted. She may wonder if they're on vacation, or if something worse happened. 

This imagined community we create in our minds through ambient familiarity can be related to Anderson's concept of the nation as an imagined community. In Anderson's structure, a nation can create an imagined community using shared media and narratives with millions of people we will never interact with \citep{anderson2006}. At the urban scale, we see that familiarity and shared spaces create an imagined community of people even if we never interact with them. 

But do we need familiarity to feel belonging? Or can we feel familiarity not only through repeated co-presence and by recognizing faces, but in other ways?

\subsection{With unfamiliar ambient co-presence}

We can feel belonging even when experiencing unfamiliar ambient co-presence events.

City-dwellers identify with their city, and imagine a community of their city. 

Let's say Sue is a New Yorker, and she's walking down 5th Avenue. Sue may not recognize a single person, but still feel a sense of belonging and community simply because the people with whom she experiences ambient co-presence share this space she calls home. 

Perhaps if everyone looks like a tourist, Sue might feel like she is not in the presence of her community. She may not feel belonging in that moment. But, of course, she will not always know for sure if every person she walks by is in fact a New Yorker. She will use her imagined concept of what her city's community is (or should be), and negotiate that with what she observes to determine if she is in the presence of her community. 

This cognitive urban community, the negotiation of that mental image with the actual individuals present, and the subsequent feeling of belonging all occur within the observer's imagination. It does not require interaction, and it does not require familiarity.

This imagination is not only active when considering a community of a city. Consider Bea sitting at the bar. Bea is Colombian, and hears a group of three speaking Spanish at the end of the bar. She overhears them talking about Calí and Bogotá. 

It is very natural for Bea's concepts of community to become activated. Maybe this bar, or maybe this city, feels a little bit more comfortable to her than it did moments ago. Maybe she realizes she yearns for a community where she can share her culture. Maybe she left Colombia for a reason and she hates being reminded of home.

Ambient co-presence can serve to activate very strong feelings of community, blending cultural and urban. But Bea might also experience much lighter touch activations to her community imagination. She overhears a patron talking to the bartender about his divorce ("he's younger than I would have expected"), she sees a married couple where one partner is in a suit and the other is in gym clothes ("did they meet here or come from home?").

In a sense, this is another form of making the unfamiliar familiar -- of breaking down stereotypes -- of filling in what was before unknown. Ambient co-presence provides data with which to make vivid your imagination of place and community. We propose here that as your imagination of community becomes more vivid, so does your ability to find a place for yourself in this imagination of community.

It is worth a moment to describe the role of ambient co-presence specifically when considering interactions across race. The Black American experience is saturated with an awareness of being noticed. Cadogan's \textit{Walking While Black} highlights many interactions, but for every interaction there are many ambient, non-interaction social experiences \citep{cadogan2016}. Even through only ambient co-presence, members of stereotyped minorities can have outsized impacts on the imaginations of others. Black Americans know this intimately.

It's important to highlight that imagined does not mean unreal. All real things we experience live in our imagination, and our imagination is a reflection of our lived reality. The key distinction is that ambient co-presence only activates, at least for one discrete event, the imagination of the observer and their concepts of community, place, and belonging. 

\section{Ambient co-presence avoids networks and closure}

Not only is imagination the main way that ambient co-presence affects observers, we can see that Granovetter's network ties are not needed to foster belonging and Coleman's network closure is not required to maintain norms in these shared urban spaces \citep{granovetter1973, coleman1988}.

Granovetter's weak ties framework requires that networks are the medium for transferring social information. But we see that belonging and sense of community can exist without the need for strong or even weak ties \citep{zahnow2025comfort, blokland2014}. These ``invisible'' or ``absent ties'' are not real ties in Granovetter's formulation, and can exist without any interaction at all. They need not ever be realized as actual network ties, and they need not be mutual. They exist in the mind of the observer, who creates these invisible ties during ambient co-presence events. The imagination (or, perhaps, potential) of ties is all that's needed to foster a sense of community.

Coleman's theory requires that norms exist when relationship networks have closure that can coordinate enforcement \citep{coleman1988}. Yet many of our norms, perhaps most of them, are learned and maintained through ambient co-presence. Simply observing the absence of a behavior acts as an implication of a norm, but that doesn't satisfy Coleman's desire to explain norm maintenance through enforcement. There's an argument to be made that norm enforcement requires occasionally breaking co-presence out of ambient mode and into engaged mode: if a norm is broken we expect one of the bystanders to break silence and use chastisement as enforcement. Perhaps there is closure in the co-presence relationships in a small urban space, but that does not seem to be the whole story.

Consider Trina on her commuter train. One day a passenger might come on playing loud music. The passenger plays their music, and nobody punishes this breach of norm directly. Yet, Trina may see the negative, but private reaction of another passenger to the loud music. That ambient co-presence event will inform Trina's concept of the norm, further reinforcing the norm in her imagined community. This does not mean that ambient co-presence can take on all the norm enforcement capabilities of a closed network of relationships, but there is certainly some capacity to impact norms through ambient observation.


% \subsection{Urban community beyond network ties}



% Granovetter's concept of network ties is useful to help describe many types of social phenomena. But when it comes to forming community and belonging, network ties do not explain the whole story.

% We do not need strong local networks to feel belonging in our neighborhood. The role of familiar strangers 


% familiar ambient co-presence




% garnette?
% gemeinschaft vs gesellschaft ?

% imagined communities / solidarity
% - familiar events create solidarity
% - being in an ethnically familiar place

% \subsection{Ambient norm enforcement}

% what about impact on norms, coleman
% - norms are enforced not through any two-sided interaction, but through a series of ambient interactions (no two people necessarily need to make eye contact for a norm to be enforced)
% - if someone is playing loud music on a train
%    - i may see others' reactions, and see that as enforcement
%    - it may bother me, and thus through ambient co-presence i derive my own categorical imperative

% The chain of ambient co-presence events

\section{Role of the shared urban space}

We have seen just a few of the ways that ambient co-presence can impact our feelings of community and belonging. And we know that ambient co-presence is pervasive activity in shared urban spaces. What role do these spaces have in impacting these social experiences and thus our feeling of community?

First, we have certainly not described the full scope of impacts that ambient co-presence can have, both on our imagination of community and on other aspects of our lives, like our mental health. There is much more we can do to understand how co-presence, both ambient and engaged, familiar and unfamiliar, conscious and unconscious, might impact how we think about ourselves, our communities, and our shared environments. It would be no surprise if having vivid, imagined feelings of community had a correlation with mental health. We know that feelings of social cohesion are also important for democracy and reducing political polarization. Of course, ambient co-presence is also one of the most prolific ways we grow our understanding of community aesthetics (e.g. fashion).

We leave our homes less than we used to \citep{bls2024}, and we also interact with other communities less \citep{surgeongeneral2023}. This is where our shared urban spaces have an opportunity to connect us with each other. If it's true that our sense of belonging can be fostered through ambient, non-interactive social experiences, then we should be designing our urban environments and planning our distribution of services, businesses, and venues accordingly.

Our urban environments should provide places where people can come repeatedly, places where they might recognize others, and where they might be recognized themselves. Even if they do not ever recognize others, we posit that even ambient co-presence around unfamiliar strangers can still be a path to stronger community.

\bibliographystyle{apalike}
\pagebreak
\bibliography{references}

\end{document}
